Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-10 paper/ch_potentials.tex
schleifen über CH Kanten vereinheitlichen
add Many-to-Many description for small k

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to target $t$}
\KwData{Minimum priority queue $Q$, also called open list, ordered by tentative distances}
$B[x] \leftarrow +\infty$ for all $x\neq t$;
$B[t] \leftarrow 0$\;
Make $Q$ only contain $t$ with weight $0$\;
\While{not $Q$ empty}{
	$y\leftarrow$ pop minimum element from $Q$\;
	\For{$xy$ is down-edge in $G^+_\ell$}{
		\If{$B[x] > w_\ell(xy) + B[y]$}{
			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
                        Add $x$ or decrease $x$'s key in $Q$ to $B[x]$\;
		}
	}
}
\caption{CH backward search}
\label{algo:ch-backward}
\end{algorithm2e}

@@

\begin{algorithm2e}
\KwData{$P[x]$: tentative distance from $x$ to $t$}
Execute Algorithm~\ref{algo:ch-backward}\;
\For{all CH levels $L$ from most to least important}{
	\For{all up edges $xy$ in $G^+_\ell$ with $x$ in $L$}{
		\If{$P[x] < P[y] + w_\ell(xy)$}{
			$P[x] \leftarrow P[y] + w_\ell(xy)$\;
		}
	}
}
\caption{PHAST basic all-to-one search}
\label{algo:phast}
\end{algorithm2e}

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		$D[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Lazy RPHAST algorithm}
\label{algo:pot}
\end{algorithm2e}

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
\KwData{$P[x]$: parent node of $x$ in the elimination tree}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
			$\Dist(P[x])$\;
		}
		$D[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Elimination tree based Lazy RPHAST algorithm}
\label{algo:cch_pot}
\end{algorithm2e}
% TODO iterative

@@

Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investigating the Lazy RPHAST performance in this setting seems worthwhile.
SSE RPHAST starts by performing the selection phase on the full target set.
% TODO flip directions?
Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
This second step can be improved by processing $k$ (typically 16) sources at once using SIMD vector instructions (SSE RPHAST).
Naively translating this approach to Lazy RPHAST would mean selecting one target after another and for each doing the full lazy exploration from all sources.
However, this approach would explore the upward search space of all the sources multiple times.
This work seems redundant.

>>>>

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to target $t$}
\KwData{Minimum priority queue $Q$, also called open list, ordered by tentative distances}
$B[x] \leftarrow +\infty$ for all $x\neq t$;
$B[t] \leftarrow 0$\;
Make $Q$ only contain $t$ with weight $0$\;
\While{not $Q$ empty}{
	$y\leftarrow$ pop minimum element from $Q$\;
	\For{all down-edges $xy$ in $G^+_\ell$}{
		\If{$B[x] > w_\ell(xy) + B[y]$}{
			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
                        Add $x$ or decrease $x$'s key in $Q$ to $B[x]$\;
		}
	}
}
\caption{CH backward search}
\label{algo:ch-backward}
\end{algorithm2e}

@@

\begin{algorithm2e}
\KwData{$P[x]$: tentative distance from $x$ to $t$}
Execute Algorithm~\ref{algo:ch-backward}\;
\For{all CH levels $L$ from most to least important}{
	\For{all up-edges $xy$ in $G^+_\ell$ with $x$ in $L$}{
		\If{$P[x] < P[y] + w_\ell(xy)$}{
			$P[x] \leftarrow P[y] + w_\ell(xy)$\;
		}
	}
}
\caption{PHAST basic all-to-one search}
\label{algo:phast}
\end{algorithm2e}

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		$D[x]\leftarrow B[x]$\;
		\For{all up-edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Lazy RPHAST algorithm}
\label{algo:pot}
\end{algorithm2e}

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
\KwData{$P[x]$: parent node of $x$ in the elimination tree}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
			$\Dist(P[x])$\;
		}
		$D[x]\leftarrow B[x]$\;
		\For{all up-edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Elimination tree based Lazy RPHAST algorithm}
\label{algo:cch_pot}
\end{algorithm2e}
% TODO iterative

@@

RPHAST was originally developed to answer many-to-many queries.
It was not designed to be used in conjunction with A*.
Investigating how Lazy RPHAST performs in the many-to-many setting is therefore interesting.
In this setting, a set of source and target nodes are given.
The objective is to compute the distances between all pairs of source and target nodes.
In this section, we described how Lazy RPHAST can be used to solve this many-to-many problem.
Our experiments will show, that the achieved performance is comparable to the original RPHAST algorithm.

To solve the many-to-many problem, RPHAST uses SIMD vector instructions.
Let $k$ denote the number of elements in one vector.
RPHAST computes the distances from all source nodes to $k$ target nodes in one run.
If the target set contains more than $k$ nodes, then the target set is split into smaller sets of $k$ elements\footnote{If the number of targets is not divisible by $k$, the set is increased by adding additional target nodes.}.
The algorithm is then run for each of these smaller sets.
Each RPHAST run performs vectorized variants of the selection and query phases.\todo{Is query phase definiert?}

We translate this setup to the Lazy RPHAST setting as follows.
First, a set of $k$ target nodes need to be fixed.
A vectorized selection phase is then performed for this fixed target set.
After the selection, the source nodes are iteratively revealed.
Once a source node is revealed, the algorithm must compute the distances to all $k$ targets.
Only after the distances are computed, the next source node is revealed.

The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:pot}.
We describe vectorized variants of these algorithms.
We replace the $B$ and $D$ arrays of scalars by arrays of $k$-dimensional vectors of scalars.
To clarify which entities are SIMD vectors, we denote the vectorized arrays by $\hat{B}$ and $\hat{D}$.
$\hat{D}[s]$ is an SIMD vector $\hat{v}$ of length $k$.
The $i$-th element $\hat{v}[i]$ of $\hat{v}$ will contain the distance from $s$ to $i$-th selected target.
We can also denote this distance using $\hat{D}[s][i]$.
All operations on SIMD vectors are component-wise.
Adding a scalar to an SIMD vector consists of adding the scalar to every element of the SIMD vector.

\todo{Rechtschreibung: down edge vs down-edge und up edge vs up-edge. Geht beides aber bitte konsistent.}

\begin{algorithm2e}
\KwIn{$\hat{T}$: SIMD vector of $k$ target nodes}
\KwData{$\hat{B}[x][i]$: tentative distance from $x$ to target $\hat{T}[i]$}
\KwData{Minimum priority queue $Q$, also called open list, ordered by node rank}
Make $Q$ an empty queue\;
$\hat{B}[x][i] \leftarrow +\infty$ for all $x$ and $i$\;
\For{all $i\in \{0\ldots k-1\}$}{
  $\hat{B}[\hat{T}[i]][i] \leftarrow 0$\;
  Add $\hat{T}[i]$ to $Q$\;
}
\While{not $Q$ empty}{
  $y\leftarrow$ pop minimum element from $Q$\;
  \For{all down-edges $xy$  in $G^+_\ell$}{
    $\hat{B}[x] \leftarrow \min\{\hat{B}[x], \hat{B}[y] + w_\ell(xy)\}$\;
    Add $x$ to $Q$ if not in $Q$\;
  }
}
\caption{Vectorized Lazy RPHAST selection}
\label{algo:simd-ch-backward}
\end{algorithm2e}

Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
In the original algorithm, the nodes are ordered by distance.
In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
To resolve this problem, we ordered the nodes in the queue not by distance but by CH rank.
The CH rank is the position of a node in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
All other modifications are straight-forward.
The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
Algorithm~\ref{algo:pot} can be vectorized by just replacing the arrays by their SIMD variants.
The result is illustrated in Algorithm~\ref{algo:simd-pot}.

\begin{algorithm2e}
\KwData{$\hat{B}[x][i]$: tentative distance from $x$ to $\hat{T}[i]$ as computed by Algorithm~\ref{algo:simd-ch-backward}}
\KwData{$\hat{D}[x][i]$: memoized distance from $x$ to $\hat{T}[i]$, $\bot$ initially}
\SetKwFunction{VecDist}{SIMDComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\VecDist{$x$}}{
  \If{$\hat{D}[x] = \bot$}{
    $\hat{D}[x]\leftarrow \hat{B}[x]$\;
    \For{all up-edges $xy$ in $G^+_\ell$}{
      $\hat{D}[x]\leftarrow\min\{\hat{D}[x],w_\ell(xy)+\VecDist(y)\}$\;
    }
  }
  \Return{$\hat{D}[x]$}\;
}
\caption{Vectorized Lazy RPHAST algorithm}
\label{algo:simd-pot}
\end{algorithm2e}

\todo{Hier beginnt der ältere Text}
Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investigating the Lazy RPHAST performance in this setting seems worthwhile.
SSE RPHAST starts by performing the selection phase on the full target set.
% TODO flip directions?
Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
This second step can be improved by processing $k$ (typically 16) sources at once using SIMD vector instructions (SSE RPHAST).
Naively translating this approach to Lazy RPHAST would mean selecting one target after another and for each doing the full lazy exploration from all sources.
However, this approach would explore the upward search space of all the sources multiple times.
This work seems redundant.
